Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-03-23 papers/wdmerger_I/paper.tex

The history of double degenerate theory can broadly be described as a
series of three major paradigm shifts. The first attempts to model the
results of the merger process came in the
1980s. \cite{nomotoiben:1985} demonstrated that off-center carbon
ignition would occur in the more massive white dwarf as it accreted
mass near the Eddington rate from the less massive white dwarf
overflowing its Roche lobe. \cite{saionomoto:1985} tracked the
evolution of the flame and found that it propagated quiescently into
the center, converting the carbon-oxygen white dwarf into an
oxygen-neon-magnesium white dwarf. This would then be followed by
collapse into a neutron star -- a result with significantly different
observational properties compared to a SN Ia. This scenario, termed
accretion-induced collapse, would be avoided only if the accretion
rate were well below the Eddington rate. \cite{tutukov_yungelson:1979}
observed that this could happen if the mass loss from the secondary
was higher than the Eddington rate and thus the accreted material
formed an accretion disc, which might rain down on the primary more
slowly. The main finding was that double degenerate systems would not
obviously lead to Type Ia supernovae.

The first three-dimensional simulations of double degenerate systems
were performed by \citet{benz:1990}, who used the smoothed particle
hydrodynamics (SPH) method to simulate the merger process. They found
that if the lower-mass star (generally called the ``secondary'') was
close enough to the more massive star (the ``primary'') to begin mass
transfer on a dynamical time scale, the secondary completely disrupted
and formed a hot envelope around the primary, with a
centrifugally-supported accretion disk surrounding the core and
envelope. They noted that the envelope temperature was hot enough to
generate carbon fusion but neglected nuclear reactions in their
simulation. Later simulations of similar setups
\citep{rasio_shapiro:1995,yoon:2007,loren-aguilar:2009,raskin:2012}
corroborated this result. This validation of the prediction of
\cite{tutukov_yungelson:1979} did not instill confidence in the
community. \cite{mochkovitch_livio:1990} and \cite{livio:2000}
observed that turbulent viscosity would be sufficiently large for
angular momentum to be removed from the disk at a rate high enough to
generate the troublesome accretion timescales. Based on all of this
evidence, the review of \cite{hillebrandtniemeyer2000} argued that the
model was only viable if the accretion-induced collapse problem could
be avoided. Later work by \cite{shen:2012} and \cite{schwab:2012} used
a more detailed treatment of the viscous transport in the outer
regions of the remnant and found that while the centrifugally
supported envelope would be converted into hot envelope material on a
viscous timescale, their simulations still led to off-center carbon
burning. \cite{vankerkwijk:2010} argued that equal-mass mergers would
lead to the conditions necessary for carbon detonation in the center
of the merged object, but \cite{shen:2012} questioned this for similar
reasons related to how viscous transport would convert rotational
motion into pressure support. \cite{zhu:2013} followed this with an
expanded parameter space study and argued that many of their
carbon-oxygen systems had the potential to detonate. The study of the
long-term evolution of the remnants is thus still an open subject of
research.

The most recent shift in perspective on this problem started with two
series of papers appearing at roughly the same time. One began with
\cite{pakmor:2010} and \cite{pakmor:2011}. This group used the SPH
method to study the merger of equal-mass ($0.9\ M_\odot$)
carbon-oxygen white dwarfs and found that a hotspot was generated near
the surface of the primary white dwarf. They argued that this region
had a temperature and density sufficient to trigger a thermonuclear
detonation. They propagated this detonation throughout the system and
found that it would observationally appear as a subluminous Type Ia
supernova. This was the first time a DD simulation successfully
reproduced at least some characteristics of a SN
Ia. \cite{pakmor:2011} tried a few different mass combinations and
found empirically that this would hold as long as the secondary was at
least 80\% as massive as the primary. These events, where the merger
process resulted in the detonation of the system during the merger
coalescence -- avoiding the much longer time-scale evolution -- were
termed ``violent'' mergers.

Around the same time, however, \cite{guillochon:2010} and
\cite{dan:2011} came to significantly different conclusions. They
pointed out that the previously mentioned simulations generally all
shared a significant drawback, which was that their initial conditions
were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006},
and \cite{motl:2007} (the first three-dimensional grid-based
simulations of mass transfer in binary white dwarf systems) pioneered
the study of looking at the long-term dynamical evolution of binary
white dwarf systems after constructing exact equilibrium initial
conditions. In contrast to the other studies, they found that paying
close attention to the accuracy of the initial conditions of the
simulation has important effects. In particular, earlier work placed
the stars too close together and ignored the effects of tidal forces
on changing the shape of the secondary, leading to the merger
happening artificially too quickly. When the initial conditions are
constructed in exact equilibrium, the system can be stable for tens of
orbital periods, substantially changing the character of the mass
transfer phase. However, one limitation of this series of studies is
that the authors used a polytropic equation of state and thus could
not consider nuclear reactions. \cite{guillochon:2010} and
\cite{dan:2011} improved on this using a realistic equation of state,
a nuclear reaction network, and a similar approach to the equilibrium
initial conditions, and found substantial agreement with the idea that
mass transfer occurs in a stable manner over tens of orbital
periods. They also found that, assuming the material accreted onto the
surface of the primary was primarily helium, explosive surface
detonations would occur as a result of accretion stream instabilities
during the mass transfer phase prior to the full merger. This could
trigger a double-detonation explosion and thus perhaps a SN Ia.

The very latest developments give some areas of agreement and some of
remaining uncertainty. \cite{pakmor:2012} performed a merger scenario
with a $1.1\ M_\odot$ and $0.9\ M_\odot$ setup, with better treatment
of the initial conditions, and also found that the merger process
happened over more than ten orbits. Nevertheless, they still found
that a carbon-oxygen detonation would occur, in line with their
earlier results. \cite{moll:2014} was also able to find a detonation
in a similarly massive system. \cite{dan:2012} and \cite{dan:2014}
performed a large sweep of the parameter space for merger pairs and
found that pure carbon-oxygen systems would generally not lead to
detonations and violent mergers except for the most massive
systems. They did find that for systems containing helium, many of
them would detonate and potentially lead to SNe Ia, either through the
aforementioned instabilities in the accretion stream, or during the
contact phase, similar to the violent carbon-oxygen
mergers. \cite{pakmor:2013} added a thin helium shell on their primary
white dwarf, and found that this robustly led to a detonation of the
white dwarf. Thus at least a preliminary agreement may be that systems
containing helium could robustly lead to events resembling SNe Ia, as
well as very massive carbon-oxygen binaries.

Given the above, why is another approach using a different simulation
code warranted? First and foremost, reproducibility of the results
across simulation codes and algorithms is important for gauging
confidence in this result. With only a couple of exceptions, most of
the existing results that study the viability of the DD progenitor as
an explanation for Type Ia supernovae (that is, including a
temperature-dependent equation of state and nuclear reactions) have
used the SPH method. SPH codes have a number of features which do aid
them in the study of these systems, such as excellent conservation of
angular momentum. However, the question of whether a prompt detonation
in a merger actually happens depends in detail on the nature of the
gas at the interface between the two stars, which is at much lower
density than the rest of the stellar material. The SPH codes use
uniform mass particles, so their effective resolution is
\textit{lowest} at the stellar surface. In contrast, a grid-based code
with adaptive mesh refinement can zoom in on the regions where
hotspots will develop, while also maintaining high enough resolution
in the high-density regions to adequately capture the large-scale mass
transfer dynamics. However, there are also outstanding questions of
convergence in SPH (e.g.\ \citealt{zhu_SPH:2014}) and whether the method
correctly captures fluid instabilities. This is an important question
for white dwarf mergers because of the likely importance small-scale
instabilities will have on the evolution of the low-density gas at the
primary's surface. The pioneering work of \cite{agertz:2007} compared
grid and SPH codes and found some important differences. Most relevant
for this discussion is that the SPH codes could not adequately handle
mixing from the Kelvin-Helmholtz instability in the test they
propose. As pointed out by \cite{price:2008}, this is not a result of
SPH being inherently unable to model this instability, but instead it
is attributed to the fact that the standard SPH evolution equations do
not have a mechanism for capturing discontinuities in internal
energy. \citeauthor{price:2008} showed that the addition of an
artificial thermal conductivity can dramatically improve the ability
of the SPH codes to exhibit this instability. There have since been a
number of other papers discussing this issue, but to our knowledge
none of these improvements have yet been incorporated into an SPH
model of a WD merger. Another reason to be skeptical is that the
detonation itself was inserted manually in \cite{pakmor:2010} based on
small-scale detonation simulations. A self-consistent evolution of a
thermonuclear detonation, if it existed, would significantly build
confidence in the progenitor model.

>>>>

The first attempts to model the results of the merger process came in the
1980s. \cite{nomotoiben:1985} demonstrated that off-center carbon
ignition would occur in the more massive white dwarf as it accreted
mass near the Eddington rate from the less massive white dwarf
overflowing its Roche lobe. \cite{saionomoto:1985} tracked the
evolution of the flame and found that it propagated quiescently into
the center, converting the carbon-oxygen white dwarf into an
oxygen-neon-magnesium white dwarf. This would then be followed by
collapse into a neutron star -- a result with significantly different
observational properties compared to a SN Ia. This scenario, termed
accretion-induced collapse, would be avoided only if the accretion
rate were well below the Eddington rate. \cite{tutukov_yungelson:1979}
observed that this could happen if the mass loss from the secondary
was higher than the Eddington rate and thus the accreted material
formed an accretion disc, which might rain down on the primary more
slowly. The main finding was that double degenerate systems would not
obviously lead to Type Ia supernovae.

Three-dimensional simulations of merging double degenerate systems were 
first performed by \citet{benz:1990}, who used the smoothed particle
hydrodynamics (SPH) method to simulate the merger process. This was 
followed later by a number of authors 
\citep{rasio_shapiro:1995,segretain:1997,guerrero:2004,yoon:2007,loren-aguilar:2009,raskin:2012}.
The main finding of these early 3D SPH simulations was that if the 
lower-mass star (generally called the ``secondary'') was
close enough to the more massive star (the ``primary'') to begin mass
transfer on a dynamical time scale, the secondary completely disrupted
and formed a hot envelope around the primary, with a
centrifugally-supported accretion disk surrounding the core and
envelope. Carbon fusion might commence in the disk, but not at a 
high enough rate to generate a nuclear detonation. \cite{mochkovitch_livio:1990} 
and \cite{livio:2000}  also observed that turbulent viscosity in this disk 
would be sufficiently large for angular momentum to be removed from the 
disk at a rate high enough to generate the troublesome accretion 
timescales discussed by \cite{tutukov_yungelson:1979}. Based on this
evidence, the review of \cite{hillebrandtniemeyer2000} argued that the
model was only viable if the accretion-induced collapse problem could
be avoided. Later work by \cite{shen:2012} and \cite{schwab:2012} used
a more detailed treatment of the viscous transport in the outer
regions of the remnant and found that viscous dissipation in the centrifugally
supported envelope would substantially heat up the envelope on a  
viscous timescale, but their simulations still led to off-center carbon
burning. \cite{vankerkwijk:2010} argued that equal-mass mergers would
lead to the conditions necessary for carbon detonation in the center
of the merged object, but \cite{shen:2012} questioned this for similar
reasons related to how viscous transport would convert rotational
motion into pressure support. \cite{zhu:2013} followed this with an
expanded parameter space study and argued that many of their
carbon-oxygen systems had the potential to detonate. The study of the
long-term evolution of the remnants is thus still an open subject of
research.

A recent shift in perspective on this problem started around 2010.
\cite{pakmor:2010} used the SPH method to study the merger of 
equal-mass ($0.9\ M_\odot$) carbon-oxygen white dwarfs and found 
that a hotspot was generated near the surface of the primary 
white dwarf. They argued that this region had a temperature 
and density sufficient to trigger a thermonuclear
detonation. They inserted a detonation which propagated throughout 
the system. They found that the result would observationally 
appear as a subluminous Type Ia supernova. This was the first time 
a DD simulation successfully reproduced at least some characteristics of a SN
Ia. \cite{pakmor:2011} tried a few different mass combinations and
found empirically that this would hold as long as the secondary was at
least 80\% as massive as the primary. These events, where the merger
process resulted in the detonation of the system during the merger
coalescence -- avoiding the much longer time-scale evolution -- were
termed ``violent'' mergers.

Around the same time, however, \cite{guillochon:2010} and
\cite{dan:2011} pointed out that the previously mentioned simulations 
generally shared a significant drawback, which was that their initial conditions
were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006},
and \cite{motl:2007} (the first three-dimensional grid-based
simulations of mass transfer in binary white dwarf systems) pioneered
the study of looking at the long-term dynamical evolution of binary
white dwarf systems after constructing exact equilibrium initial
conditions. Earlier work placed the stars too close together 
and ignored the effects of tidal forces that change the shape of the 
secondary, leading to the merger
happening artificially too quickly. When the initial conditions are
constructed in exact equilibrium, the system can be stable for tens of
orbital periods, substantially changing the character of the mass
transfer phase. However, one limitation of this series of studies is
that the authors used a polytropic equation of state and thus could
not consider nuclear reactions. \cite{guillochon:2010} and
\cite{dan:2011} improved on this using a realistic equation of state,
a nuclear reaction network, and a similar approach to the equilibrium
initial conditions, and found substantial agreement with the idea that
mass transfer occurs in a stable manner over tens of orbital
periods. They also found that, assuming the material accreted onto the
surface of the primary was primarily helium, explosive surface
detonations would occur as a result of accretion stream instabilities
during the mass transfer phase prior to the full merger. This could
trigger a double-detonation explosion and thus perhaps a SN Ia.

The latest developments have resulted in some possible areas of convergence.
\cite{pakmor:2012} performed a merger scenario
with a $1.1\ M_\odot$ and $0.9\ M_\odot$ setup, with better treatment
of the initial conditions, and indeed found that the merger process
happened over more than ten orbits. Nevertheless, they still determined
that a carbon-oxygen detonation would occur, in line with their
earlier results. \cite{moll:2014} was also able to find a detonation
in a similarly massive system. Very recently, \cite{kashyap:2015} 
found that a detonation occurred in a similarly massive system. Notably,
the detonation occurred self-consistently and did not need to be inserted 
by hand. \cite{dan:2012} and \cite{dan:2014} performed a large sweep 
of the parameter space for merger pairs and
found that pure carbon-oxygen systems would generally not lead to
detonations and violent mergers except for the most massive
systems. They did find that for systems containing helium, many
would detonate and potentially lead to SNe Ia, either through the
aforementioned instabilities in the accretion stream, or during the
contact phase, similar to the violent carbon-oxygen
mergers. \cite{pakmor:2013} added a thin helium shell on their primary
white dwarf, and found that this robustly led to a detonation of the
white dwarf. For now there is preliminary support for the hypothesis
that systems with helium shells (or helium WDs), and very massive carbon-oxygen binaries,
could robustly lead to events resembling SNe Ia.

Given the above, why is another approach using a different simulation
code warranted? First and foremost, reproducibility of the results
across simulation codes and algorithms is important for gauging
confidence in this result. Most of the existing results that study 
the viability of double degenerate systems as progenitors for
Type Ia supernovae (that is, including a realistic 
equation of state and nuclear reactions) have
used the SPH method. SPH codes have a number of features which do aid
them in the study of these systems, such as excellent conservation of
angular momentum. However, the question of whether a prompt detonation
in a merger actually happens depends in detail on the nature of the
gas at the interface between the two stars, which is at much lower
density than the rest of the stellar material. The SPH codes use
uniform mass particles, so their effective resolution is
\textit{lowest} at the stellar surface. In contrast, a grid-based code
with adaptive mesh refinement can zoom in on the regions where
hotspots will develop, while also maintaining high enough resolution
in the high-density regions to adequately capture the large-scale mass
transfer dynamics. However, there are also outstanding questions of
convergence in SPH (e.g.\ \citealt{zhu_SPH:2014}) and whether the method
correctly captures fluid instabilities. This is an important question
for white dwarf mergers because of the likely importance small-scale
instabilities will have on the evolution of the low-density gas at the
primary's surface. The pioneering work of \cite{agertz:2007} compared
grid and SPH codes and found some important differences. Most relevant
for this discussion is that the SPH codes could not adequately handle
mixing from the Kelvin-Helmholtz instability in the test they
propose. As pointed out by \cite{price:2008}, this is not a result of
SPH being inherently unable to model this instability, but instead it
is attributed to the fact that the standard SPH evolution equations do
not have a mechanism for capturing discontinuities in internal
energy. \citeauthor{price:2008} showed that the addition of an
artificial thermal conductivity can dramatically improve the ability
of the SPH codes to exhibit this instability. There have since been a
number of other papers discussing this issue, but to our knowledge
none of these improvements have yet been incorporated into an SPH
model of a WD merger. Another reason for caution is that other than the
most recent results of \cite{kashyap:2015}, no simulation has self-consistently
resulted in a thermonuclear detonation. Reproducibility of the detonation 
through direct numerical simulation is critical for building 
confidence in this progenitor model.
